---
title: Reporting of Results for Dissertation
---

## basic AIPW 
```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(spatstat)

win <- owin(c(-1, 1), c(-1, 1))

beta <-t( t(rnorm(6)))
treatment_link <- function(x,y) cbind(1, x, y, x^2, y^2, x * y) %*% beta

treatment_mu <- function(x,y) plogis(treatment_link(x,y))

treatment <- function(x,y) rbinom(length(x), 1, prob = treatment_mu(x,y))

beta2 <- t(t(rnorm(6)))
confound <- function(x,y) cbind(1, x, y, x^2, y^2, x * y) %*% beta2

treatment <- as.function(as.im(treatment, win))
confound <- as.function(as.im(confound, win))

p <- process <- rpoispp(function(x,y) exp(2 + 2 * treatment(x,y) + confound(x,y)), win = win)


tt <- as.im(treatment)

for_reg <- mutate(expand_grid(x = tt$xcol, y = tt$yrow), v = c(tt$v))

prop_fit <- glm(v ~ poly(x,y, degree = 2), data = for_reg, family = binomial())

prop <- predict(prop_fit, type = 'response')
vv <- tt$v

vv[] <- prop
propensity_score <- tt
propensity_score$v <- vv
outcome <- ppm(p ~ treatment +poly(x,y, degree = 2))


tt_new = as.im(1, win)
new <- list(treatment = tt_new)
mu <- predict(outcome, type = 'intensity', covariates = new)
new <- list(treatment = tt_new - 1)
mu2 <- predict(outcome, type = 'intensity', covariates = new)

Y <- as.im(p)
a <- c(tt$v)

## looks correct
## first term is integration with respect to intensity measure, i.e. the expectaiton.
second term is integration with respect to residual measure, i.e. 
y <- integral(mu) + sum(tt * Y / propensity_score) - integral(tt * mu / propensity_score)
y2 <- integral(mu2) + sum((1-tt) * Y / (1-propensity_score)) - integral((1-tt) * mu2 / (1-propensity_score))
```



```{r}
## basic composition

library(tidyverse)
library(tidycensus)
library(sf)

d <- read_csv('~/Desktop/decennial_race_data/DECENNIALPL2020.P1_data_with_overlays_2021-12-20T114314.csv', skip = 1)
dr <- d %>% select(1:2, contains('Population of one race')) 

dr <- dr %>% select(!3)


library(compositions)

dr %>% pack(comp = c(3:8)) %>%
    mutate(comp = ilr(comp))
bg <- tigris::block_groups(state = 'PA')

bg <- bg %>% transmute(id = GEOID)
dr <- dr %>% mutate(id = str_extract(id, '(?<=US).+'))

dr <- dr %>% select(1, 2, 3, 4, 6, 8) %>% mutate(across(!c(1:2), ~ . + ))

dr <- dr %>% pack(comp = (3:6)) %>%
    mutate(comp = ilr(comp))
dr <- dr %>% mutate(comp = as.data.frame(comp)) %>%
    unpack(cols = comp)
dr <- dr %>% select(-2)

data <- inner_join(dr, bg)
data <- st_as_sf(data)

dd <- dist(select(as_tibble(data), V1, V2, V3))

dd <- as.matrix(dd)
example <- 1

datat <- data %>% mutate(dist_from = c(dd[example,]))

point <- st_centroid(data[example,])

ggplot() + scale_fill_viridis_c() +
    geom_sf(aes(fill = dist_from), data = datat) +
    geom_sf(data = point, size = 2)



## basic raster fits

library(tidyverse)
library(stars)
library(partykit)
d <- read_ncdf('~/Downloads/V5GL02.HybridPM25.NorthAmerica.201907-201907.nc')

d <- filter(d, lon > -91, lon < -90, lat < 41, lat > 40)

r <- d$GWRPM25
y <- c(r)
co <- st_coordinates(d)
dd <- bind_cols(co, y = y)
library(mgcv)
fit <- gam(y ~ s(lon, lat, bs = 'gp', m = 5), data = dd)
r2y <- predict(fit)
r2 <- r
r2[!is.nan(r2)] <- r2y

par(mfrow = c(3,1))
image(r)
image(r2)
image(r - r2)

fit <- lmtree(y ~ lon + lat | lon + lat, data = dd)
p <- predict(fit, newdata = dd)

r3 <- r2
r3[] <- p
image(r3)

resid <- r - r3

fit <- gam(y ~ s(lon, lat, bs = 'gp', m = 5), data = mutate(dd, y = c(resid)))

p2 <- predict(fit, newdata = dd)


library(rSPDE)
library(INLA)

## loc_2d_mesh = co
## mesh_2d = inla.mesh.2d(
##   loc=loc_2d_mesh,
##   cutoff=0.05,
##   max.edge=c(0.1,0.5) )


## Abar <- rspde.make.A(mesh = mesh_2d, loc = as.matrix(loc_2d_mesh))
## mesh.index <- rspde.make.index(name = 'field', mesh = mesh_2d)

## st.dat <- inla.stack(
##   data = list(y = as.vector(r)),
##   A = Abar,
##   effects = mesh.index)


## rspde_model <- rspde.matern(mesh = mesh_2d, nu_upper_bound = 1)


## f = y ~ -1 + f(field, model = rspde_model)
## ## inla currently not working
## ## rspde_fit <- inla(f, data = inla.stack.data(st.dat),
##                   ## family = 'gaussian',
##                   ## control.predictor = list(A = inla.stack.A(st.dat)),
## ## inla.mode = 'experimental', verbose = TRUE)

mlik <- function(theta, Y, A, op) {
  sigma = exp(theta[1])
  kappa = exp(theta[2])
  nu = exp(theta[3])
  return(-rSPDE::rSPDE.matern.loglike(object = op,
                                      Y = Y,
                                      A = A,
                                      user_kappa = kappa,
                                      user_sigma = sigma,
                                      user_nu = nu, sigma. = exp(theta[4])))
}

loc_2d_mesh = co[!is.nan(y), ]
y <- y[!is.nan(y)]
mesh_2d = inla.mesh.2d(
  loc=loc_2d_mesh,
  cutoff=0.05,
  max.edge=c(0.1,0.5) )


library(optimParallel)
n_cores <- parallel::detectCores() - 1
cl <- makeCluster(n_cores)

setDefaultCluster(cl = cl)
sigma <- 1
range <- 0.2
nu <- 0.5
kappa <- sqrt(8*nu)/range
op <- matern.operators(mesh=mesh_2d,nu=nu,
                                   kappa=kappa,sigma=sigma,m=2)

theta0 = c(get.inital.values.rSPDE(mesh = mesh_2d),
           log(0.1 * sqrt(var(as.vector(y), na.rm = TRUE))))

A <- inla.spde.make.A(
  mesh=mesh_2d,
  loc=as.matrix(loc_2d_mesh))

sigma.e <- 0.1
pars <- optimParallel(theta0, mlik, Y = y, A = A, op = op)
results <- data.frame(sigma = exp(pars$par[1]), 
                      kappa =  exp(pars$par[2]),
                      nu = exp(pars$par[3]),
                      sigma.e = exp(pars$par[4]),
                      row.names = "Estimate")


predgrid <- inla.mesh.projector(mesh_2d, xlim = c(0,1), 
ylim = c(0,1))
A.prd2 <- predgrid$proj$A


sigma.e.est <- exp(pars$par[4])
op.prd <- update(op, user_sigma = exp(pars$par[1]),
                 user_kappa = exp(pars$par[2]),
                 user_nu = exp(pars$par[3]))

pred.rspde <- predict(op.prd, A = A, Aprd = A, Y = y, 
                      sigma.e = sigma.e.est,
                      compute.variances = TRUE)
rp <- r
rp[!is.nan(r)] <- pred.rspde$mean

```



```{r}

```



True DGP Illustration

```{r}
library(tidyverse)
library(splines)
set.seed(42)
t <- seq(from = 0, to =  1, length.out = 5000)
x <-  ns(t, df = 15) %*% (rnorm(15))
y <-  ns(t, df = 7) %*% rexp(7)


dat <- tibble(t, x, y) |>
    mutate(
        roughx = x + rnorm(1000, 0, 0.05),
        roughy = y + rnorm(1000, 0, 0.05))

## Motivating plot: suppose that the "truth" that generates real-life data is a complicated mathematical structure (not even a function necessarily).
## in the motivating figure we think of the truth being the black curve and the noisy path around representing the "data".


ggplot(dat, aes(x,y)) +
    geom_path() +
    geom_path(aes(roughx, roughy), color = 'gray', alpha = 0.4) +
    geom_point(aes(roughx, roughy), size = 0.01, alpha = 0.1) + theme_void()


py <- predict(lm(y ~ ns(t, 8), data = dat))
px <- predict(lm(x ~ ns(t, 8), data = dat))


mutate(dat, py = py, px = px) |>
    ggplot( aes(x,y)) +
    geom_path() +
    geom_path(data = dat, aes(px, py), color = 'red', alpha = 0.4) +
    geom_point(aes(roughx, roughy), size = 0.01, alpha = 0.1) + theme_void()


```



## scritp-to-vignette

library(tidyverse)
library(rmarkdown)
library(knitr)
## library(onion)
library(spatstat)

clause_length <- 100
nclause <- 100
k <- clause_length * nclause
lits <- rpois(k, lambda = 73)
pars <- rbinom(n =k, size = 1, prob = 0.5)
## use the pars

## because we're just generating random witnesses we don't have to even bother
## actually doing the permutation, just generate the random witness and then index by names.
## If we count the number of names first we can speed this up further, and rename as well,
## I won't bother to do that here though


groups <- rep(1:nclause, each = clause_length)
random_sequence <- function(clause_length, nclause)
{
    k <- clause_length * nclause
    ## let F be an arbitrary distribution for generating literals

    witness <- rbinom(n = k, size = 1, prob = 0.5)

    assignment <- map_int(split(witness[lits], groups), sum)

    logistic_dv <- cumand(if_else(assignment > 0, 1, 0))

    condition on the number of times lit <x> occurs
    condition on the number of times  a given clause is satisfied.


}

## string valued expression


## Point process in a box?
## pick a compact window with intensity 1, we'll adjust the window dimension to change the effective intensity of the points.

## We'll superimpose point processes based on interaction thresholds.
## We'll impose bell's inequality hard treshold for the interaction, but approximate from below with a b-spline approximation to a sigmoid that collapses to a step function asymptotically. We'll use tanh theoretically and b-splines for approximation. We'll actually also sequentially increase a negative penalty on the derivative one step higher than the function we are trying to estimate directly (maybe the other way?)

## How about sequential alpha-shape/voronoi union and radon transform estimation?
## I'll use rgb as well to capture the 1 interaction, 2-interaction, and 3-interaction processes via marking point processes.

## bxd <- c(0,1)
## bb <- boxx(bxd, bxd, bxd, bxd)

## h <- rpoisppx(10, bb)

## ; y^2 = x^3 + ax + b
## cubes assumed to start in one corner and follow arbitrary convention of directions.
cube <- function(...) {
    x <- c(...)
    class(x) <- 'compact'
    x
}

## assumed to have origin at 0
ball <- function(eps) {

}








## basic idea here is fractional roots probably are fractal objects in nature.
## maybe the other way around here
## maybe also fractal dimension?
## So a 2/3 root is probably a 2-dimensional fractal object in 3-dimensional space.
## and a 3/2 root is probably a set of 3-dimensional fractial objects's intersection with 2-dimensional space


## let's just hamiltonian estimate this stuff, or maybe just poisson simulate, we'll see.

## Oh yeah let's follow t at some set of points and restrict ball size progressively.

## (x + yi)

xi <- rnorm(1000)
yi <- rnorm(1000)




## next up convolve this with a gaussian approximation
a <- 2
b <- 0.5

rnorm(1000 * 7)

## iterated regression projection here
## 8-dimensional t that is 7+1
## hmm if we interpret only 1 as time we can sort it iteratively, maybe independently.



## x <- tcone %*% beta + rnorm(length(t))
## y <-  (x^2 + a * x + b + tcone %*% beta + rnorm(length(t)))^(1/3)



## okay new goal is to iteratively get an estimate in high dimensional space of a surface that basically just intersects 2d space at the elliptic curve.
## oh, so why not just make that a penalty term instead?
## honestly makes perfect sense

## this will be a nice way to estimate elliptic integrals.
## also a nice way to learn how to like estimate some quantum shit

## library(mgcv)
## ym <- gam(y ~ te(tcone, k = 12))
## xm <- gam(x ~ te(tcone, k = 12))



## xp <- predict(xm, newdata = list(tcone = tcone))
## yp <- predict(ym, newdata = list(tcone = tcone))
## tm <- gam(t ~ te(x,y, k = 12))

satval <-function(x, ...){
    UseMethod("satval")
}

or_clause <- function(x1, x2, x3) {
    l <- list(x1, x2, x3)
    class(l) <-'or_clause'
    return(l)
}

or_vec <- function(x) {
    or_clause(x[1], x[2], x[3])
}


and_clause <- function(x1, x2){
    l <- list(x1, x2)
    class(l) <-'and_clause'
    return(l)
}

sat_form <- function(Aclause, litstring, m = '', n = '') {
    s <- list(form = Aclause, litstring = litstring, m = m, n = n)
    class(s) <-'sat_form'
    return(s)
}


## x will be a number. If positive treat as literal.
## If negative treat as not-expression
format.literal <- function(x){
    if(x > 0){ paste0('x', x)}
    else if (x < 0){ paste0('¬', 'x', abs(x))}
    else {'XX'}
}

print.or_clause <- function(x, ...){
    x <- lapply(x, format.literal)
    s <- paste0('(', paste0( x, collapse = ' ∨ '), ')')
    cat(s)
    cat('\n')
    invisible(s)
}

print.and_clause <- function(x, ...){
    x <- lapply(x, \(x) capture.output(print(x)))
    cat(paste0(x, collapse = ' ∧ '))
    cat('\n')
}


print.sat_form <- function(x, ...){
   print(x$form)
}
## in R we will assume we will be passed symbols in the form of numeric literals.
## We will use negative to indicate nots.
## However we will pretty print everything for human readability.

## input_expression is a list of non 0 valued integers that represent literals.
## We will assume this is perfectly parseable linearly by preprocessing elsewhere
test_exp <- 1:(3*10)


accumulate_lits <- function(lits, clause_size) {
    mclause <- length(lits) %/% clause_size

    index <- 0
    for (m in 1:mclause) {
        terms <- rep(NA, times = clause_size)
        for (i in 1:clause_size) {
            index <- index + 1
            terms[i] <- lits[index]
        }
        ## construct or_clause
        clause <-or_clause(terms[[1]], terms[[2]], terms[[3]])
        if (m == 1){
            form <- clause
        } else {
            form <- and_clause(form, clause)
        }
    }
    form
}

start_frame <- function(nvec) {
    n <- length(nvec)
    m <- (n %/% 3)

    lit_frame <- tibble(lit = nvec) |>
        mutate(
            clause_id = rep(1:m[[1]], each = 3),
            )

    clause_frame <- select(lit_frame, clause_id) |>
        mutate(clause = accumulate_lits(lit_frame$lits, m))

    list(lit_frame = lit_frame, clause_frame = clause_frame)
}


random_expression <-function(m, lambda, clause_size) {
    n <- clause_size * m
    signature <- if_else(rbinom(n, size = 1, prob = 0.50) == 1, 1L, -1L)
    names <- rpois(n, lambda)

    unique_names <- length(unique(names))
    assignment <- signature * names
    form <- accumulate_lits(assignment)


    ## everything basically just for pretty printing up to here
    sat_form(form,  litstring = assignment, m = m, n = unique_names)
}




## okay do this streaming
## much easier that way and will prevent memory issues

## so we're going to generate randomly.
## then we'll rename this random list to sorted integers
## then use those to reconstruct the original statement

## then evaluate that statement
## at a random witness

rename <- function(ex){
    ## rename from 1 up for each new number
    ## if we encounter something we've seen before we should rename it according to past ruleset
    int <- 1:length(ex)
    o <- order(abs(ex))
    exo <- ex[o]
    shuffle <- int[o]
    shuffle[o]

}

monte_step <- function(n){
}

    ## precondition requirement is that this is an exact multiple of 3, which it should be since this is 3-sat



## radon transform

## splines


## hitting functions

## point processes


## percolator


## abm

## montecarlo runtime


## re-attempt downhere
n <- 30


rrademacher <- function(n, p) {
    if_else(rbinom(n, 1, p) ==1, 1, -1)
}

rename_succ <- function(x){
    retvec <- 0 * x
    for (i in seq_along(x)){
        if (i == 1) {retvec[i] <- 1}
        else if( x[i] != x[i-1]){ retvec[i] <- retvec[i-1] + 1}
        else {retvec[i] <- retvec[i-1]}
    }
retvec
}


## basic monte carlo assessment program

## now add error handling, pop it on a trampoline, etc
## we'll migrate this into racket now to make it easier to study, but make calls to Rcmd for statistical calls

## regression strategy
## group by literal name and perform exact logistic regression strategy.
## Let literals be the
monte <- function(){
    set.seed(42)
    n <- 30
    init <- tibble(id = 1:n) |>
        mutate(string = rpois(n, 5)) |>
        group_by(string) |>
        mutate(rename = min(id)) |>
        arrange(rename) |>
        ungroup() |>
        mutate(
            sign = rrademacher(n, 8/9),
            rename = rename_succ(rename),
            nlit = max(rename))

    or_frame <- mutate(init,
                       group_clause = rep(1:(n/3), each = 3)) |>
        group_by(group_clause) |>
        summarize(or_clause = list(or_vec(rename * sign )))

    wac <- reduce(or_frame$or_clause, and_clause)

    nlit <- init$nlit[1]
    witframe <- tibble(rename = 1:nlit, assignment = rbinom(nlit, 1, 8/9))

    evaluator_frame <- left_join(init, witframe)


    evaluator <- function(assignment, sign){
        evaluation <- if_else(sign ==1,
                              assignment,
                              1 - assignment)
    }

    valuation <- function(evaluator_frame) {
        with(evaluator_frame,
             assignment,  sign)
    }

    satisfaction <- function(valuation) {
        groups <- rep(1:(length(valuation)/3), each = 3)

        reduce(lapply(split(valuation, groups), sum), prod)
    }

    satisfaction(valuation(evaluator_frame))
}




## let's try setting the poisson rate parameter to log(length) of the string, then I bet we can show with probability 1 that every sentence of length 1 will be represented in the output of this process in polynomial time.


  library(readr)
  library(sf)
  map <- read_rds('R/simulation_utilities/allegheny_region_tracts.rds')

## sample code for generating the fem object
library(tidyverse)
topo <- map |> st_make_valid() |> 
  group_split(id) |> map(st_union) |> 
  map_dfr(st_as_sf)

t1 <- st_as_sfc(topo[1,])

## pa coords--40.4406° N, 79.9959° W
t1 <- st_transform(t1, "EPSG:3857") |> st_simplify(dTolerance = 200)

library(fmesher)
# create mesh
edge <- st_buffer(t1, 3000)
me <- fm_mesh_2d(boundary = edge, interior = t1, max.edge = 2500)
library(INLA)

Amat <- inla.spde.make.A(me)
#fm_basis

spde_mod <- inla.spde2.matern(me, alpha = 2)





#compositional covars





## kernel_sim
source('outcome_models.R')

library(spatstat)

x <- swedishpines

kern <- function(x,y) exp(-5 * sqrt((x^2 + y^2)))

xe <- as.im(x)

ts <- seq(from = -1, to = 1, length.out = 256)

t2 <- seq(from = -2, to = 2, length.out = 256 * 2)
km <- (outer(t2, t2, kern))

zero_pad <- matrix(0, nrow = 256 * 2, ncol = 256 * 2)



f1 <- zero_pad
f1[1:256, 1:256] <- as.matrix(as.im(x, dimyx = 256))
f1 <- fft(f1)

f2 <- fft(apply(apply(km, 1, fftshift), 1, fftshift))


resurf <- fft(f1 * f2, inverse = TRUE) / prod(dim(f1))
resurf <- Re(resurf)[1:256, 1:256]
image(resurf)



y <- c(resurf) + rnorm(length(resurf))

ym <- matrix(y, nrow = 256, 256)
yf <- as.function.im(as.im(ym, window(swedishpines)))
Q <- quadscheme(swedishpines)

dat <- mpl_prepare(swedishpines, quadscheme(swedishpines), ppcov = list(x), dimyx = c(128, 128))

qc <- coords(Q)
yy <- yf(qc[,1], qc[,2])

dat$yy <- yy
fit <- gam(yy ~ s(`...5`, bs = 'conv') - 1, data = dat)

basis <- smoothCon(s(`...5`, bs = 'conv'), data = dat, absorb.cons = TRUE)
basis <- basis[[1]]$X

dat$fitted <- fitted(fit)

smooths <- fit$smooth

smooth_internal <- smooths[[1]]$internal_basis

XX <- PredictMat(smooth_internal, data.frame(distances = seq(from = 0, to = 3, length.out = 400)))

plot(XX[,2:10] %*% coef(fit))





## simulation.R

## generative model candidate library set
## it would be very useful to generalize the model to an entire set of candidate models, retain information about which ones are consistent with the simulation scenario and then prune incompatible models.

## And it would also be nice to be able to set a fixed database where we can use a bayesian bootstrap to approximate the desired model by sampling from that database rather than necessarily having to regenerate everything every single time, at least within a desired level of aproximation accuracy.

## assumptions
positivity: P(A) > 0, P(!A) > 0
estimability: A _|_ X | X, f(U) where f ~ Class
exchangeability: A _|_ Y | X
consistency: Y(A = 1) = Y(1)


## a simulation is an entire data analysis that CAN BE REPEATED arbitrarily many times
simulation <- function( ) {

}


experiment <- function() {}

estimator <- function() {}

estimand <- function() {}

distribution <- function(){}


## simulation.Rmd

---
output: html_document
---
```{r, include = FALSE}
library(RandomFields)
library(tidyverse)
library(sf)
library(spatstat)
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE)
library(disslib)
```

#####################
## BASIC DATA PREP ##
#####################

Sim 1
```{r}
## for all simulations we will use the basic map of allegheny county
allegheny <- read_rds('simulation_utilities/allegheny_region_tracts.rds') |>
    mutate(region = id, id = row_number())
allegheny <- st_make_valid(allegheny)
allegheny <- st_transform(allegheny, '+proj=aeqd +lon_0=-79.3212891 +lat_0=40.2041883 +datum=WGS84 +units=m +no_def')


ggplot(allegheny) + geom_sf(aes( fill = region))

covering_grid <- st_make_grid(allegheny, n = c(20, 20))
gg <- st_intersection(covering_grid, allegheny)


alltess <- map(allegheny$geometry, as.owin)
alltess <- tess(tiles = alltess)

window <- as.owin(allegheny)

## generate point process valued outcome with confounding by covariates for basic binary exposure.
## for simulation 1 we consider a simple, spatially located binary treatment with a non-confounding covariate


alltessf <- as.function(alltess)
set.seed(12)

treatment_center <- runifpoint(1, window)
treat_dist <- distfun(treatment_center)

treatmentfun <- function(x,y) treat_dist(x,y) < 12000
treatment <- as.im.function(treatmentfun, window)


covar_values <- rnorm(nrow(allegheny))
covarfun <-function(x,y) covar_values[as.numeric(alltessf(x,y))]
covar <- as.im.function(covarfun, window)


## the intercept term is specified as the expected number of points with treatment and covariate set to 0
## to get a reasonable and roughly controllable process it is best to set this term manually as log(N /area(w))
## where N is the expected number of points for the "neutral" process and area is the area of the observation window
```

Sim 1

```{r, sim1}
#######################
## BINARY TREATMENTS ##
#######################

intercept <- log(100 / area(window))
intens <- function(x,y) exp(covarfun(x,y) * 2 + treatmentfun(x,y) + intercept)
intens_im <- as.im(intens, window)
pp <- rpoispp(intens_im)

ppm(pp ~ covar + treatment)
ppm(pp ~ covar + treatment + s(x,y, k = 40, bs = 'gp'), use.gam = TRUE)

ppm(pp ~ treatment + s(x,y, k = 40, bs = 'gp'), use.gam = TRUE)
glmdata <- getglmdata(ppm(pp ~ alltessf + treatment))

ff <- glm(.mpl.Y ~ treatment + alltessf, weights = .mpl.W, data = glmdata, family = poisson())

plot(treatment)
```

Sim 2

```{r, sim2}
## simulation 2
## pointwise binary exposure, including covariates, but not confounding
## pointwise on the fine 128 by 128 grid used for computation
## technically this is still blockwise, but for small blocks

treatment2 <- rbinom(128 * 128, size = 1, prob = 0.5)
treatment2 <- matrix(treatment2, 128, 128)
treatment2 <- as.im(treatment2, window)

plot(treatment2)

intens <- exp(intercept + treatment2 + covar)
pp <- rpoispp(intens)
ppm(pp ~ treatment2 + covar)
```

Sim 3

```{r, sim3}
## simulation3
## binary random field exposure
## generated from matern random field and then clipped on the basis of being greater or lower than zero
gridx <- seq(from = 0, to = 1, length.out = 128)
treatment3 <- as.im(as.matrix(RFsimulate(RMmatern(var = 2, scale = 0.3, nu = 1.3), gridx, gridx)) > 0, window)

plot(treatment3)

intens <- exp(intercept + treatment3 + 0.1 * covar)

plot(intens)
pp <- rpoispp(intens)
fit <- ppm(pp ~ treatment3 + covar, use.gam = TRUE)
fit <- ppm(pp ~ treatment3 + covar + s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)
```

Sim 4
```{r}
## simulation 4, binary exposure with lower degree of spatial autocorrelation
treatment4 <- as.im(as.matrix(RFsimulate(RMmatern(var = 2, scale = 0.05, nu = 1.3), gridx, gridx)) > 0, window)

plot(treatment4)

intens <- exp(intercept + treatment4 + 0.1 * covar)
pp <- rpoispp(intens)
fit <- ppm(pp ~ treatment4 + covar, use.gam = TRUE)
fit <- ppm(pp ~ treatment4 + covar + s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)
dd <- getglmdata(fit)
fit <- gam(.mpl.Y ~ treatment4 + covar + te(x,y, bs = 'gp', k = 10), weights = .mpl.W, data = dd, family = poisson())


treatment5 <- as.im(as.matrix(RFsimulate(RMmatern(var = 2, scale = 0.001, nu = 1.3), gridx, gridx)) > 0, window)

intens <- exp(intercept + treatment5 + 0.1 * covar)
pp <- rpoispp(intens)
fit <- ppm(pp ~ treatment5 + covar, use.gam = TRUE)
fit <- ppm(pp ~ treatment5 + covar + s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)
dd <- getglmdata(fit)
fit <- gam(.mpl.Y ~ treatment5 + covar + te(x,y, bs = 'gp', k = 10), weights = .mpl.W, data = dd, family = poisson())
```
## simulation structure

Sim 5
```{r}
## next round of simulations
## consider scenarios where covariates are themselves spatially autocorrelated, so there is residual spatially determined confounding.
spatial_covar <- as.im(as.matrix(RFsimulate(RMmatern(var = 2, scale = 0.1, nu = 1.3), gridx, gridx)), window)
intens <- exp(intercept + treatment3 + spatial_covar)



pp <- rpoispp(intens)
fit <- ppm(pp ~ treatment3 + spatial_covar, use.gam = TRUE)
fit <- ppm(pp ~ treatment3)
fit <- ppm(pp ~ treatment3 +  s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)


```

Sim 6

```{r}
## explicitly confounded treatment on the basis of spatially autocorrelated covariate
treatment6 <- as.im(matrix(rbinom(128^2, 1, plogis(c(as.matrix(spatial_covar)) * 0.3 - 1)), 128, 128), window)

plot(treatment6)

plot(spatial_covar)

intens <- exp(intercept + treatment6 + spatial_covar)
pp <- rpoispp(intens)

fit <- ppm(pp ~ treatment6,  use.gam = TRUE)
fit <- ppm(pp ~ treatment6 +  spatial_covar,  use.gam = TRUE)
fit <- ppm(pp ~ treatment6 +  s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)
```

Sim 7

```{r}
## treatment 7 is a spatially autocorrelated treatment resulting counfounded by a spatially autocorrelated covariate,
## but that is realized as a direct sum of the covariate and a secondary random process.
## as a result, this treatment has additional spatial autocorrelation and is 'clumpy'
residual_process <- RFsimulate(RMmatern(1.2, var = 0.8, scale = 0.7), gridx, gridx)
treatment7t <- as.im(as.matrix(residual_process), window) + spatial_covar
treatment7t <- as.function(treatment7t)
treatment7 <- function(x,y) treatment7t(x,y) >0
treatment7 <- as.im(treatment7, window)

plot(treatment7)
plot(spatial_covar)


intens <- exp(intercept + treatment7 + spatial_covar)
pp <- rpoispp(intens)

fit <- ppm(pp ~ treatment7,  use.gam = TRUE)
fit <- ppm(pp ~ treatment7 +  spatial_covar,  use.gam = TRUE)
fit <- ppm(pp ~ treatment7 +  s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)

```

Sim 8

```{r}

## treatment8
## generate treatment as set of circles where circle centers are generated from a poisson process on the basis of the confounding covariate
set.seed(0)
pp_base <- spatial_covar - min(spatial_covar) + 1

## generate approximately 10 exposure centers
pp_base <- 10 * pp_base / integral(pp_base)

centers <- rpoispp(pp_base)

treat_surf <- distfun(centers)

treatment8 <- as.im(as.function(function(x,y) treat_surf(x,y) < 5000), window)

plot(treatment8)

intens <- exp(intercept + treatment8 + spatial_covar)
pp <- rpoispp(intens)

fit <- ppm(pp ~ treatment8,  use.gam = TRUE)
fit <- ppm(pp ~ treatment8 +  spatial_covar,  use.gam = TRUE)
fit <- ppm(pp ~ treatment8 +  s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)
```
###########################
## CONTINUOUS TREATMENTS ##
###########################

```{r}
## treatment is spatial covar plus a residual surface

treatment9 <- as.im(treatment7t, window)

residual_process_noisy <- RFsimulate(RMmatern(1.2, var = 0.8, scale = 0.7) + RMnugget(var = 0.3), gridx, gridx)
treatment10t <- as.im(as.matrix(residual_process_noisy), window) + spatial_covar
plot(as.im(treatment10t, window))
plot(treatment9)
intens <- exp(intercept + treatment9 + spatial_covar)
pp <- rpoispp(intens)

fit <- ppm(pp ~ treatment9,  use.gam = TRUE)
fit <- ppm(pp ~ treatment9 +  spatial_covar,  use.gam = TRUE)
fit <- ppm(pp ~ treatment9 +  s(x,y, bs = 'gp', k = 40),  use.gam = TRUE)


## realistic and potentially advantageous scenario
## we have a continuous treatment generated as a function of space, but spatial confounders have discontinuities due to geographical features
sim_surface <- function(){
    im <- as.function(as.im(as.matrix(RFsimulate(RMmatern(nu = 1, var = 2, scale = 1), gridx, gridx))))

}
s <- map(1:4, ~sim_surface())
## determine data generation mechanism
## pick optimal model from candidate model set according to specified criteria
## calculate truth value of specified estimand under dgp
## generate data
## calculate estimator values on data

## aggregate estimator values
## calculate summary statistics on estimators
## combine summary statistics across simulations for comparisons

indexer <- function(x,y) allegheny$region[as.numeric(alltessf(x,y))]
indexed <- as.im(indexer, window)

partials <- map(1:4, function(i)as.im(function(x,y) s[[i]](x,y) * (indexer(x,y) == i), window))
broken_covar <- reduce(partials, `+`)

plot(broken_covar)

```


```{r}


set.seed(34)
source('outcome_models.R')

library(tidyverse)
library(spatstat)
## swedishpines <- murchison$gold

## swedishpines <- shift(swedishpines, -c(352782.9, 6699742))


## it appears there's some sort of interpolation bug due to window scale going on that results in covariates not scaling properly.
## may also have something to do with distance calculations
myexposure <- rescale(swedishpines, c(10))
## myexposure <- redwood
## myexposure <- rescale(redwood, 0.67)

## myexposure <- swedishpines
exp_im <- (as.matrix(as.im(myexposure)))
kernx <- seq(from = -12, to = 12, length = 128 * 2)
kern <- fft2shift(outer(kernx, kernx, function(x,y) exp(- x ^2 - y^2)))

exp_imb <- matrix(0, nrow = 256, ncol = 256)
exp_imb[1:128, 1:128] <- exp_im


exp_im <- Re((fft(fft(exp_imb) * fft(kern), inverse = TRUE) / 256^2))[1:128, 1:128]
## intens_surface <- as.im(exp_im / 100, W = myexposure$window)
intens_surface <- as.im(exp_im * 10, W = myexposure$window)



## dd <- dd |> mutate(mytreatment = treatcov)
make_summary_plot <- function(model, intens, newpoints){

    
    plot_data <- mutate(newpoints,
                        pred = predict(model, newpoints, type = 'response'),
                        truth = interp.im(intens, newpoints),
                        res = truth - pred
                        ) 

    plot_data |>
        pivot_longer(c(truth, res, pred), names_to = 'plot', values_to = 'value') |>
        ggplot(aes(x, y, fill = value)) + geom_raster() + facet_wrap(~plot) + scale_fill_viridis_c()
}




mytreatment <- (intens_surface > median(intens_surface))



treated_intens_surface <- exp(log(intens_surface))
myoutcome <- (rpoispp(lambda = treated_intens_surface))
nd <- 200
q <- quadscheme(myoutcome, nd = nd)
mydummy <- as_tibble(coords(q$dummy))
mydd <- mydummy |>
    slice(-((nrow(mydummy) - 3):nrow(mydummy)))



fit2 <- ppmod(myoutcome, q, ppcov = list(myexposure = myexposure), k = 40)
mydd2 <- mutate(mydd, mytreatment = mytreatment[list(x = x, y = y)])
make_summary_plot(fit2, treated_intens_surface, mydd2)


## show visual outcome model in basic model, including other covariates
# outcome model estimation




```

Emigration Sim--collab with Dr. Marsh


```{r load_libraries, include = FALSE}
library(tidyverse)
library(janitor)
library(readxl)
library(mipfp)
library(DT)

set.seed(12)
```
## Methods

### Rates
Rates were taken from the white male rates
file for England from the chloroprene study. All age groups were used from the last time period available
(1996-2000). These rates were taken to be constant over time for all
time periods, and were applied for both men and women in the cohort. 

```{r read_death_rates, include = FALSE, cache = TRUE}
start <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 110)
end   <- c(18, 28, 38, 48, 58, 68, 78, 88, 98, 108, 118)
names <- c('46-50', '51-55', '56-60', '61-65', '66-70', '71-75', '76-80', '81-85', '86-90', '91-96', '96-00')

age_groups <- c('< 5',  '5-9', '10-14', '15-19', '20-24', '25-29', '30-34', '35-39', '40-44', '45-49', '50-54', '55-59', '60-64', '65-69', '70-74', '75-79', '80-84', '85+')

col_positions <- fwf_positions(start = start, end = end, col_names = names)
## read in rates for latest time period and use only this by default
rates <- read_fwf('england_rates.txt', skip = 3, col_positions) %>%
    mutate(age_group = age_groups) %>%
    gather(time_period, rate, -age_group) %>%
    filter(time_period == '96-00') %>%
    select(-time_period)

## age bounds for rates for non-interval valued age ranges
rates_extreme <- filter(rates, age_group %in% c('< 5', '85+')) %>%
    mutate(lb = c(0, 85), ub = c(5, Inf)) %>%
    select(-age_group)

## age bounds for rates with interval valued age ranges
rates_normal <- filter(rates, !(age_group %in% c('< 5', '85+'))) %>%
    separate(age_group, into = c('lb', 'ub'), sep = '-') %>%
    mutate_at(c('lb', 'ub'), as.integer)


## final set of rates
rates <- bind_rows(rates_extreme, rates_normal) %>%
    select(lb, ub, rate)


```

```{asis, include = params$verbose_output}
Below is the rates table used here. The the variable "lb" is the age lower bound and "ub"
the age upper bound. These represent lower and upper (inclusive)
    bounds for the various age groups.
```

```{r, echo = FALSE, include = params$verbose_output}
datatable(rates)
```

### Cohort replication
As person year data from the original chloroprene cohort was not
immediately available, it was necessary to recreate a simulated cohort
from the available tables in the first chloroprene paper. Constraints
that were useful for recreating sex and age data were entered into the
program and used to constrain the cohort simulation.

```{r import_constraints_and_tables, include = FALSE}
# file where chloroprene tables as well as immigration table were manually entered
source('table2.R')
## import tables into local environment
list2env(tables, envir = environment())
```

In order to simulate the starting cohort in the year 2000, the
following variables were used from the chloroprene paper: age at hire,
time since first employment, duration of employment, vital status, and
sex. The available counts for these variables were used to fill in the
margins of a multidimensional contingency table. A multidimensional
iterative proportional fitting procedure was then used to obtain cell
weight estimates for every cell in the contingency table. Cell weights
were integerized with the truncate-replicate-sample algorithm. This
algorithm was used only once to obtain an initial cohort for
simulation. The algorithm is subject to slight randomness, meaning
this introduces additional estimation uncertainty that has not yet
been accounted for. 

After simulating a cohort, an initial cohort for simulation was
obtained by removing all dead individuals from the sampled cohort. Age
at hire and time since first employment data were used to obtain age
group upper bounds and lower bounds. These were restricted to reasonable
width by assuming no one is hired before the age of 16, no one is
hired after the age of 60, and that no one was initially employed more
than 35 years ago. The
cohort was further simplified in order to reduce the complexity of the
simulation by assuming everyone was male. 

```{r cohort_microsim, include = FALSE, cache = TRUE}
## the truncate replicate sample algorithm
trs <- function(table) {
    counts <- table$Freq
    rounded <- floor(counts)
    decimal <- counts - rounded
    counts <- as.integer(counts)
    topup <- sample(length(counts), round(sum(decimal)), prob = decimal)
    counts[topup] <- counts[topup] + 1
    table$integer_counts <- counts
    table
}


## use the constraint tables from chloroprene paper
cons <- constraints
cons <- cons[ c('age_at_hire', 'time_since_first_employment', 'duration_of_employment', 'vital_status', 'sex')]

vitalc <- cons$vital_status
vitalc <- with(vitalc, list(alive = alive_confirmed + alive_assumed, dead = dead_unknown + dead_known))
cons$vital_status <- vitalc

cons <- map(cons, ~unlist(unlist(.)))

seedt <- array(1, dim = map(cons, length), dimnames = map(cons, ~names(.)))
target.list <- 1:5
weights <- Ipfp(seedt, target.list, cons)

## get weights from ipfp algorithm
df <- as.data.frame.table(weights[[1]]) %>% as_tibble()
## sample the cohort
df <- trs(df)


## clean up the cohort
cohort <- df %>% select(-Freq, -duration_of_employment) %>%
    filter(vital_status == 'alive') %>%
    select(-vital_status) %>%
    ## assume for simplicity that lowest hiring age possible is 16 and oldest is 60
    ## additionally assume that longest time since employment possible is 35 years
    mutate(
        age_at_hire = fct_recode(age_at_hire, '16-19' = '<20', '30-60' = '30+'),
        time_since_first_employment = fct_recode(time_since_first_employment, '0-20' = '<20', '30-35' = '30+')
    ) %>%
    separate(age_at_hire, into = c('aah_lb', 'aah_ub'), sep = '-') %>%
    separate(time_since_first_employment, into = c('tsfe_lb', 'tsfe_ub'), sep = '-') %>%
    mutate_at(vars(aah_lb, aah_ub, tsfe_lb, tsfe_ub), parse_integer) %>%
    mutate(
        age_lb = tsfe_lb + aah_lb,
        age_ub = tsfe_ub + aah_ub
    ) %>%
    select(age_lb, age_ub, sex, integer_counts)

## aggregate the cohort over sex
cohort <- cohort %>%
    group_by(age_lb, age_ub) %>%
    summarize(n = sum(integer_counts)) %>%
    ungroup()

```

```{asis, include = params$verbose_output}
The cohort contains a total of `r sum(cohort$n)` living individuals
    out of `r sum(cons[[1]])` from the original cohort.
```

```{r, echo = FALSE, include = params$verbose_output}
datatable(cohort)
```

## Population Data

Annual mid-year population estimates were obtained from Northern
Ireland official data. Full country population estimates were available by
sex and age. The official data contained single year population
estimates, so population numbers were aggregated into age groups to
match those available for emigration data. Population counts were
available from 2000 to 2017. Final population
counts were then aggregated across both male and female to get age
group and year specific population counts. 

```{r population_data, include = FALSE, cache = TRUE}
## 2000 to 2017 mid year estimates in northern ireland by age
pop_data <- read_excel('MYE17_SYA.xlsx', sheet = 2) %>%
    filter(
        str_detect(area,'Northern Ireland'),
        gender %in% c('Males', 'Females'),
        year %in% 2000:2017
    ) %>%
    select(age, year, MYE, gender)

## aggregate pop data into age groups
## age groups are selected to be the same as the emigration data
pop_data <- pop_data %>%
    mutate(age_group = case_when(
               age < 18 ~ 'Less than 18 years',
               age %in% 18:24 ~ '18-24',
               age %in% 25:34 ~ '25-34',
               age %in% 35:44 ~ '35-44',
               age %in% 45:54 ~ '45-54',
               age %in% 55:64 ~ '55-64',
               age >= 65 ~ '65 years and over'
           ),
           ## rename genders for matching
           gender = case_when(
               gender == 'Females' ~ 'Female',
               gender == 'Males' ~ 'Male'
           )) %>%
    group_by(age_group, gender, year) %>%
    summarize(n_pop = sum(MYE))

## Extrapolate population data by directly carrying forward counts
pop_data_extrap <- pop_data %>% filter(year == 2017)
pop_data_extrap <- bind_rows(
    pop_data_extrap %>% mutate(year = 2018),
    pop_data_extrap %>% mutate(year = 2019),
    pop_data_extrap %>% mutate(year = 2020))

## add extrapolated data into the true data
pop_data <- pop_data %>% bind_rows(pop_data_extrap)

```

```{r, echo = FALSE, include = params$verbose_output}
datatable(pop_data)
```

## Emigration Data

Emigration data was made available in terms of total emigration
counts nation-wide for the years 2007 through 2017 in terms of both
age group and sex. Since emigration counts were need for the full range from 2000 to 2016, extrapolation by year
was performed with a Poisson regression model with year, gender, and
age group as covariates. Expected emigration counts were then
computed for each year for which extrapolation was necessary. 
Total emigration counts were aggregated across sex
to obtain year and age group specific counts. The poisson regression
model was taken as-is and was not further validated nor was any
further model selection carried out. 

The resulting emigration count dataset was merged with population by
year and by age group in order to compute later required rates.

```{r immigration_data, include = FALSE, cache = TRUE}
# age group emigration numbers
age_group <- age_group %>%
    clean_names() %>%
    ## the next two lines carry gender data down and remove the aggregated gender counts
    mutate(gender = if_else(row_number() <= 8, 'Male', 'Female')) %>%
    slice(-c(1, 9)) %>%
    gather(time_period, n, -gender_age, -gender) %>%
    rename(age = gender_age, n_age = n) %>%
    mutate(year = str_sub(time_period, 5, 8), year = parse_integer(year), time_period = NULL)


## fit a poisson model on age and time in order to extrapolate immigration counts
## to years where data is not available (2000-2006, and 2018-2020)
fit.extrap <- glm(n_age ~ gender + age + year, data = age_group, family = 'poisson')

## create dummy dataset for extrapolating
extrap_age <- age_group %>% filter(year == 2007)
age_group_extrap <- map_dfr(c(2000:2006, 2018:2020), ~mutate(extrap_age, year = .))

## create counts for unknown years by generating predictions from poisson model
pred <- broom::augment(fit.extrap, newdata = age_group_extrap, type.predict = 'response') %>%
    select(age, gender, year, n_age = .fitted) %>%
    mutate(n_age = ceiling(n_age))

## add simulated years back to actual years
emigration <- bind_rows(pred, age_group) %>% arrange(year)

## join emigration data to population data
emigration <- emigration %>%
    rename(age_group = age) %>%
    left_join(pop_data) %>%
    ungroup() %>%
    group_by(age_group, year) %>%
    summarize(n_age = sum(n_age),
              n_pop = sum(n_pop))

## get the upper bounds and lower bounds on the non interval age ranges
emigration_extreme <- filter(emigration, age_group %in% c('Less than 18 years', '65 years and over')) %>%
    ungroup() %>%
    mutate(lb = case_when(
               age_group == 'Less than 18 years' ~ 0,
               age_group == '65 years and over' ~ 65
           ),
           ub = case_when(
               age_group == 'Less than 18 years' ~ 17,
               age_group == '65 years and over' ~ Inf
           )) %>%
    select(-age_group)

## upper and lower bounds on interval age ranges
emigration_normal <- filter(emigration, !(age_group %in% c('Less than 18 years', '65 years and over'))) %>%
    ungroup() %>%
    separate(age_group, into = c('lb', 'ub'), sep = '-') %>%
    mutate_at(c('lb', 'ub'), as.integer)

## final emigration counts dataset
emigration <- bind_rows(emigration_extreme, emigration_normal)

```

```{r, echo = FALSE, include = params$verbose_output}
datatable(emigration %>% select(lb, ub, year, n_emigrate = n_age, n_pop) %>% arrange(lb))
```


### Simulation

The simulation was conducted by aging the cohort one year at a time
from the year 2001 to the year 2016. In each year emigration and
deaths were simulated by generating poisson random variables according
to each age group's rates and current counts. Simulated emigrants and
deaths were subsequently subtrated from the cohort, and then the
cohort aged. The simulation was run with two strategies: all
emigration simulated first and all deaths simulated first. Both
simulation strategies were run for 1,000 repetitions. 

Death rates and emigration rates were complicated by non-overlapping
age ranges. In order to select death rates, the narrowest set of age
ranges which entirely contained the age range of each age group in the
cohort was
selected. Rates were averaged across the selected age ranges and the
new average rate was applied to all individuals in the respective
cohort group. For emigration rates a similar procedure was
utilized for determining relevant age groups, although
rates were also determined on the basis of year. Emigration rates for
the selected age ranges were then
directly computed based on emigration and population counts. 

```{r simulation_functions, include = FALSE, cache = TRUE}
age <- function(data) {
    data %>% mutate(age_lb = age_lb + 1,
                    age_ub = age_ub + 1)
}

choose_rates <- function(age_lb, age_ub) {
    rates_to_apply <- filter(rates, age_lb <= ub, lb <= age_ub)
    ## annual rate per 1k
    rate_to_apply <- max(rates_to_apply$rate)
    rate_to_apply / 1000
}
choose_emigration <- function(age_lb, age_ub, year_cohort) {
    emigration_to_apply <- filter(emigration, year == year_cohort, age_lb <= ub, lb <= age_ub)
    emigration_rate <- sum(emigration_to_apply$n_age) / sum(emigration_to_apply$n_pop)
    emigration_rate
}
die <- function(data) {
    ## determine_rates
    local_rates <- map2_dbl(data$age_lb, data$age_ub, ~choose_rates(.x, .y))
    counts <- map_int(data$n, ~rpois(1, local_rates  * .))
    list(total_count = sum(counts), cohort = mutate(data, n = n - counts))
}

emigrate <- function(data, year) {
    local_rates <- pmap_dbl(list(data$age_lb, data$age_ub, rep(year, nrow(data))), ~choose_emigration(..1, ..2, ..3))
    counts <- map_int(data$n, ~rpois(1, local_rates * .))
    list(total_count = sum(counts), cohort = mutate(data, n = n - counts))
}

```

## Results

```{r run_simulation, include = FALSE}

starting_cohort <- cohort
years <- 2001:2016
die_first <- function(starting_cohort) {
    cohort <- starting_cohort
    emigrant_cohort <- cohort %>% mutate(n = 0)
    
    yearly_death_counts <- rep(0, length(years))
    yearly_emigrant_death_counts <- rep(0, length(years))
    yearly_emigration_counts <- rep(0, length(years))
    ## cohort starts at 2000
    for (i in seq_along(years)) {
        ## deaths
        death_sim <- die(cohort)
        death_em_sim <- die(emigrant_cohort)
        yearly_death_counts[i] <- death_sim$total_count
        yearly_emigrant_death_counts[i] <- death_em_sim$total_count
        
        cohort <- death_sim$cohort
        emigration_cohort <- death_em_sim$cohort
        ## emigration
        em_sim <- emigrate(cohort, years[i])
        emigrant_cohort <- mutate(emigrant_cohort, n = n + (cohort$n - em_sim$cohort$n))
        cohort <- em_sim$cohort

        yearly_emigration_counts[i] <- em_sim$total_count
        ## age the cohort
        cohort <- age(cohort)
    }
    list(dead = sum(yearly_death_counts) + sum(yearly_emigrant_death_counts),
         emigrated = sum(yearly_emigration_counts),
         dead_after_emigration = sum(yearly_emigrant_death_counts))
}

die_first_results <- map_dfr(1:1000, ~die_first(starting_cohort))

emigrate_first <- function(starting_cohort) {
    cohort <- starting_cohort
    emigrant_cohort <- starting_cohort %>% mutate(n = 0)
    yearly_death_counts <- rep(0, length(years))
    yearly_emigrant_death_counts <- rep(0, length(years))
    yearly_emigration_counts <- rep(0, length(years))
    ## cohort starts at 2000
    for (i in seq_along(years)) {
        ## emigration
        em_sim <- emigrate(cohort, years[i])
        emigrant_cohort <- mutate(emigrant_cohort, n = n + (cohort$n - em_sim$cohort$n))
        cohort <- em_sim$cohort

        yearly_emigration_counts[i] <- em_sim$total_count
        ## deaths
        death_sim <- die(cohort)
        death_em_sim <- die(emigrant_cohort)
        yearly_death_counts[i] <- death_sim$total_count
        yearly_emigrant_death_counts[i] <- death_em_sim$total_count
        
        cohort <- death_sim$cohort
        emigration_cohort <- death_em_sim$cohort
        ## age the cohort
        cohort <- age(cohort)
    }
    list(dead = sum(yearly_death_counts) + sum(yearly_emigrant_death_counts),
         emigrated = sum(yearly_emigration_counts),
         dead_after_emigration = sum(yearly_emigrant_death_counts))
}

emigrate_first_results <- map_dfr(1:1000, ~emigrate_first(starting_cohort))
```

Final results for the two simulation strategies are presented in the table
below. Namely, mean number of total deaths, emigrants and number of
emigrants who later died across the cohorts
over the fifteen year simulation time span, as well as quantiles for
each quantity. The variable "dead" includes all individuals who died,
including those who emigrated and later died. The variable "emigrated" includes the
total number of people who ever emigrated. The variable "dead after
emigration" includes all individuals who died after emigrating.

```{r results, include = FALSE, cache = TRUE}
starting_n <- sum(starting_cohort$n)
percent_of_cohort <- function(col) map_chr(col, ~glue::glue(round(100 * . / starting_n, digits= 2), '%'))
```

### Results from Death First Simulation
```{r, echo = FALSE}
die_first_results %>% gather(status, value) %>%
    group_by(status) %>%
    summarize(mean = mean(value),
              `2.5%` = quantile(value, 0.025),
              `97.5%` = quantile(value, 0.975)) %>%
    mutate(mean = str_c(mean, ' (', percent_of_cohort(mean), ')'),
           `2.5%` = str_c(`2.5%`, ' (', percent_of_cohort(`2.5%`), ')'),
           `97.5%` = str_c(`97.5%`, ' (', percent_of_cohort(`97.5%`), ')')) %>%
    knitr::kable()

```

### Results from Emigration First Simulation
```{r, echo = FALSE}
emigrate_first_results %>% gather(status, value) %>%
    group_by(status) %>%
    summarize(mean = mean(value),
              `2.5%` = quantile(value, 0.025),
              `97.5%` = quantile(value, 0.975)) %>%
    mutate(mean = str_c(mean, ' (', percent_of_cohort(mean), ')'),
           `2.5%` = str_c(`2.5%`, ' (', percent_of_cohort(`2.5%`), ')'),
           `97.5%` = str_c(`97.5%`, ' (', percent_of_cohort(`97.5%`), ')')) %>%
    knitr::kable()
```


```{asis, include = params$verbose_output}
Here is a sample simulation person-counts table. Age groups are listed
as the initial age group. Each year column is the total count of
people remaining in the cohort (neither died nor emigrated) at the end
of that year. 
2000 similar tables are generated in the course of the simulations. A second
set of similar tables (not shown) are created in which emigrants
accumulate and then die at the same rates as the general cohort.
```


### Sample persons-table
```{r , cache = TRUE, echo = FALSE, include = params$verbose_output}
years <- 2001:2016
die_first_s <- function(starting_cohort) {
    cohort <- starting_cohort
    cohorts <- map(1:length(years), ~starting_cohort)
    ## cohort starts at 2000
    for (i in seq_along(years)) {
        ## deaths
        death_sim <- die(cohort)
        cohort <- death_sim$cohort
        ## emigration
        em_sim <- emigrate(cohort, years[i])
        cohort <- em_sim$cohort
        cohorts[[i]] <- cohort
        ## age the cohort
        cohort <- age(cohort)
    }
    cohorts
}

cohorts <- die_first_s(starting_cohort)
initial_age_groups <- starting_cohort %>% unite(age_group, age_lb, age_ub, sep = '-') %>% pull(age_group)
initial_n <- starting_cohort %>% pull(n)
cohorts <- map2_dfr(cohorts, years, ~mutate(.x, year = .y))
cohorts %>% mutate(initial_age_groups = rep(initial_age_groups, length(years))) %>%
    select(-age_lb, -age_ub) %>%
    spread(year, n) %>%
    mutate(initial_n = initial_n) %>%
    select(initial_age_groups, initial_n, everything()) %>%
    knitr::kable()
```



CMR Code



library(tidyverse)

set.seed(11)

plants <- 1:8

## create_person
n <- 35000

birth_years <- 1960:1990

rates <- read_csv('rates.csv') |>
    janitor::clean_names() |>
    select(contains('label'), 'population', 'deaths') |>
    rename_with(~str_remove(.x, '_label')) |>
    mutate(age_group_start = parse_integer(str_sub(age_group_12, 1, 2)),
           time_period_start = parse_integer(str_sub(time_period, 1, 4)),
           rate = deaths / population) |>
    select(-c(`age_group_12`, time_period, population, deaths)) |>
    rename(race = white_non_white, cod = cod_category_code) |>
    arrange(race, sex, age_group_start, time_period_start) |>
    mutate(rate = if_else(is.na(rate), 0.0, rate))

allcause <- rates |> filter(cod == 'All Causes of Death') |>
    select(!cod)

age_groups <- unique(allcause$age_group_start)
periods <- unique(allcause$time_period_start)

indexer <- function(age, year, race, sex){
    index <- 1 + (race == 'White') * 312 +
        (sex == 'Male') * 156 +
        (max(which(age_groups <= age)) - 1) * 13 + 
        (max(which(periods <= year))) - 1
    index
}

sample_death_time <- function(start_age, start_year, race, sex){
    force_death <- 110
    time_death <- 0


    while(time_death < force_death){
        i <- indexer(start_age, start_year, race, sex)
        time_death_temp <- time_death + max(rexp(allcause[1, i]))
                                            
    }
}
                           
    
demo_table <- tibble(
    plant = sample(plants, size = n, replace = TRUE),
    sex = sample(c('M', 'F'), size = n, replace = TRUE, prob = c(0.75, 0.25)),
    race = sample(c('B', 'W'), size = n, replace = TRUE, prob = c(0.40, 0.60)),
    smoker = sample(c(TRUE, FALSE), size = n, replace = TRUE, prob = c(0.90, 0.1)),
    wage = sample(c('BC', 'WC'), size = n, replace = TRUE, prob = c(0.90, 0.1)),
    birth_year = sample(birth_years, size = n, replace = TRUE),
    birth_month = sample(1:12, size = n, replace = TRUE)) |>
    mutate(
        hire_month = sample(1:12, size = n, replace = TRUE),
        hire_year = sample(18:27, size = n, replace = TRUE) + birth_year,
        retire_year = sample(64:68, size = n, replace = TRUE),
        retire_month = sample(1:12, size = n, replace = TRUE))
    
    

## race

## sex


library(tidyverse)
library(popEpi)
library(lubridate)
library(furrr)
library(ggforce)
plan(multiprocess)

N <- 10000
set.seed(860)


## WRITE TESTS

## this is slow but not the bottleneck code
rpwe <- function(pyt) {
    rate <- pyt$true_rate
    times <- pyt$per
    for (i in 1:(length(rate) - 1)) {
        sample <- times[i] + rexp(1, rate = rate[i])
        if (sample <= times[i + 1]) return(sample)
    }
    rexp(1, rate[length(times)]) + times[length(times)]
}

cumhaz_single <- function(t, pyt) {


    rates <- pyt$rate
    pyrs <- pyt$pyrs
    per <- pyt$per

    if (t <= per[1]) return(0)
    
    i <- 1
    cumhaz <- 0

    for (i in 1:(length(per))) {
        if (i == length(per)) {
            cumhaz <- cumhaz + rates[i] * (t - per[i])
        }
        else if (t > per[i + 1]) {
            cumhaz <- cumhaz + rates[i] * pyrs[i]
        }
        else {
            cumhaz <- cumhaz + rates[i] * (t - per[i])
            break
        }

    }
    cumhaz
}


## write unit tests for this including the cases where active index has to be found!!!!
survival <- function(t, pyt) {
    stopifnot(sort(t) == t)
    cumhaz <- vector('double', length(t))

    rates <- pyt$rate
    pyrs <- pyt$pyrs
    per <- pyt$per

    active_index <- 1

    for (i in 1:(length(per))) {
        if (i == length(per) & t[1] >= per[i]) {
            active_index <- i
        }
        else if (per[i] <= t[1] & per[i + 1] >= t[1]) {
            active_index <- i
            break
        }
    }
    
    cumhaz[1] <- cumhaz_single(t[1], pyt)

    if (length(t) == 1) {
        return(exp(-cumhaz))
    }
    
    for (i in 2:length(t)) {
        ## if (active_index == 1 & t[i] <= per[1]) {
        if (t[i] <= per[1]) {
            cumhaz[i] <- 0
        } else if (active_index == length(per)) {
            cumhaz[i] <- cumhaz[i-1] + rates[active_index] * (t[i] - t[i-1])
        }
        else if (t[i] >= per[active_index + 1]){
            cumhaz[i] <- cumhaz[i-1] +
                rates[active_index] * (per[active_index + 1] - t[i-1]) +
                rates[active_index + 1] * (t[i] - per[active_index + 1])
            active_index <- active_index + 1
        }
        else {
            cumhaz[i] <- cumhaz[i-1] + rates[active_index] * (t[i] - t[i-1])
        }
    }

    exp(-cumhaz)
}

person_years_single <- function(t, pyt) {


    rates <- pyt$rate
    pyrs <- pyt$pyrs
    per <- pyt$per

    if (t <= per[1]) return(0)
    
    i <- 1
    cumhaz <- 0
    person_years <- 0

    for (i in 1:(length(per))) {
        if (i == length(per)) {
            cumhaz <- cumhaz + rates[i] * (t - per[i])
            person_years <- person_years + exp(-cumhaz) * (t - per[i])
        }
        else if (t > per[i + 1]) {
            cumhaz <- cumhaz + rates[i] * pyrs[i]
            person_years <- person_years + exp(-cumhaz) * pyrs[i]
        }
        else {
            cumhaz <- cumhaz + rates[i] * (t - per[i])
            person_years <- person_years + exp(-cumhaz) * (t - per[i])
            break
        }

    }
    person_years
}



## this has a bug!!!!!
expected_person_years <- function(t, pyt) {
    stopifnot(sort(t) == t)
    cumhaz <- vector('double', length(t))
    person_years <- vector('double', length(t))

    rates <- pyt$rate
    pyrs <- pyt$pyrs
    per <- pyt$per

    active_index <- 1

    for (i in 1:(length(per))) {
        if (i == length(per) & t[1] >= per[i]) {
            active_index <- i
        }
        else if (per[i] <= t[1] & per[i + 1] >= t[1]) {
            active_index <- i
            break
        }
    }
    
    cumhaz[1] <- cumhaz_single(t[1], pyt)
    person_years[1] <- person_years_single(t[1], pyt)
    if (length(t) == 1) {
        return(person_years)
    }
    
    for (i in 2:length(t)) {
        ## if (active_index == 1 & t[i] <= per[1]) {
        if (t[i] <= per[1]) {
            cumhaz[i] <- 0
            person_years[i] <- 0
        } else if (active_index == length(per)) {
            cumhaz[i] <- cumhaz[i-1] + rates[active_index] * (t[i] - t[i-1])
            person_years[i] <- person_years[i-1] + exp(-cumhaz[i]) * (t[i] - t[i-1])
        }
        else if (t[i] >= per[active_index + 1]){
            cumhaz[i] <- cumhaz[i-1] +
                rates[active_index] * (per[active_index + 1] - t[i-1]) +
                rates[active_index + 1] * (t[i] - per[active_index + 1])
            person_years[i] <- person_years[i-1] + exp(-cumhaz[i]) * ((per[active_index + 1] - t[i-1]) + (t[i] - per[active_index + 1]))
            active_index <- active_index + 1
        }
        else {
            cumhaz[i] <- cumhaz[i-1] + rates[active_index] * (t[i] - t[i-1])
            person_years[i] <- person_years[i-1] + exp(-cumhaz[i]) * (t[i] - t[i-1])
        }
    }

    person_years
}

expected2 <- function(t, pyt) -log(survival(t, pyt))
expected <- function(t, pyt) 1 - survival(t, pyt)


## this code belongs below
observed <- function(sim, eval_times) { map_int(eval_times, ~sum(sim$sim <= .)) }

observed_matrix <- function(sim, eval_times) {
    mat <- sim %>%
        mutate(dead = map(sim, ~eval_times > .)) %>%
        pull(dead)
    mat <- do.call(rbind, mat)
    colnames(mat) <- eval_times
    rownames(mat) <- sim$id
    mat
}

time_matrix <- function(sim, eval_times) {
    mat <- do.call(rbind, map(1:nrow(sim), ~eval_times))
    colnames(mat) <- eval_times
    rownames(mat) <- sim$id
    mat
    
}

age_matrix <- function(sim, eval_times) {
    mat <- sim %>%
        mutate(age_at_time = map(birth, ~eval_times - decimal_date(.))) %>%
        pull(age_at_time)
    mat <- do.call(rbind, mat)
    colnames(mat) <- eval_times
    rownames(mat) <- sim$id
    mat
}

fot_matrix <- function(sim, eval_times) {
    mat <- sim %>%
        mutate(fot = map(entry, ~eval_times - decimal_date(.))) %>%
        pull(fot)
    mat <- do.call(rbind, mat)
    colnames(mat) <- eval_times
    rownames(mat) <- sim$id
    mat
}

cmr_expected <- function(simulation, eval_times) {
    expected_deaths <- rep(0, length(eval_times))
    for (i in 1:nrow(simulation)) {
        current <- simulation[i,]
        expected_deaths <- expected_deaths + expected(eval_times, (current$pyt)[[1]])
    }
    expected_deaths
}

stratify <- function(sim, eval_times, mat_func, lowers, uppers) {
    stopifnot(length(lowers) == length(uppers))
    oc <- o <- observed_matrix(sim, eval_times)
    cmrc <- cmr <- cmr_expected_matrix(sim, eval_times)
    smrc <- smr <- smr_expected_matrix(sim, eval_times)
    constraint_matrix <- mat_func(sim, eval_times)

    out <- vector('list', length(lowers))

    for (i in 1:length(lowers)) {
        constraints <- constraint_matrix <= uppers[i] & constraint_matrix >= lowers[i]

        oc[] <- as.integer(if_else(constraints, o, FALSE))
        cmrc[] <- if_else(constraints, cmr, 0)
        smrc[] <- if_else(constraints, smr, 0)

        cmr_computed <- colSums(oc) / colSums(cmrc)
        smr_computed <- colSums(oc) / colSums(smrc)

        out[[i]] <- list(cmr = cmr_computed, smr = smr_computed, lower = lowers[i], upper = uppers[i])
    }
    out
    
}


cmr_expected_matrix <- function(simulation, eval_times) {
    expected_lists <- map(1:nrow(simulation), ~double(length = length(eval_times)))
    expected_deaths <- rep(0, length(eval_times))
    for (i in 1:nrow(simulation)) {
        current <- simulation[i,]
        expected_lists[[i]] <- expected(eval_times, (current$pyt)[[1]])
    }
    mat <- do.call(rbind, expected_lists)
    colnames(mat) <- eval_times
    rownames(mat) <- simulation$id
    mat
}

## crude code for computing stratified smrs
## plerp  <-simu %>% mutate(age_at_time = map(en, ~ eval_times - .))
## simu %>% mutate(age_at_time = map(en, ~ eval_times - .)) %>% pull(age_at_time) %>% {do.call(rbind, .)} -> age_mat
## cmr_expected_mat <- do.call(rbind, cmr_expected_list(simu, eval_times))
## cmr_expected_mat[!(age_mat >= 20 & age_mat <= 40)] <- 0
## o <- imap_int(eval_times, function(time, index) sum(plerp$sim <= time &
                                             ## map_dbl(plerp$age_at_time, ~pluck(., index)) <=40 &
                                             ## map_dbl(plerp$age_at_time, ~pluck(., index)) >= 20))

## not correct currently
smr_expected <- function(simulation, eval_time) {

    expected_deaths <- 0
    
    for (i in 1:nrow(simulation)) {
        current <- simulation[i,]
        death_time <- current$sim
        
        expected_deaths <- expected_deaths + expected2(pmin(eval_time, death_time) , (current$pyt)[[1]])
    }
    expected_deaths
}

smr_expected_matrix <- function(simulation, eval_times) {
    expected_lists <- map(1:nrow(simulation), ~double(length = length(eval_times)))
    expected_deaths <- rep(0, length(eval_times))
    for (i in 1:nrow(simulation)) {
        current <- simulation[i,]
        death_time <- current$sim
        
        expected_lists[[i]] <- expected2(pmin(eval_times, death_time), (current$pyt)[[1]])
    }
    mat <- do.call(rbind, expected_lists)
    colnames(mat) <- eval_times
    rownames(mat) <- simulation$id
    mat
}





rates <- read_csv('alleghe_rates.csv') %>%
    transmute(race = str_to_lower(White_NonWhite_Label),
              sex = str_to_lower(Sex_Label),
              age = Age_Group_12_Label %>% str_sub(1, 2) %>% parse_integer(),
              time = Time_Period_Label %>% str_sub(1, 4) %>% parse_integer(),
              rate = Rate / 1000) %>%
    ## start intially by using time == 1950 for everyone for simplicity
    group_by(race, sex, time) %>%
    mutate(
        start_age = age, end_age = lead(age, default = Inf)) %>%
    filter(start_age >= 10)





rate_ages <- unique(rates$age)
rate_pers <- unique(rates$time)

init_cohort <- tibble(race = c(rep('white', times = 4000), rep('nonwhite', 1000)),
                      sex = c(rep('male', 2000), rep('female', 2000), rep('male', 500), rep('female', 500)),
                      entry_age = rdunif(5000, 20, 50),
                      entry_year = rdunif(5000, 1950, 1990),
                      entry_time = runif(5000, 0, 1)) %>%
    mutate(entry = date_decimal(entry_year + entry_time),
           birth = date_decimal(decimal_date(entry) - entry_age),
           exit = as_date(date_decimal(3000)),
           id = row_number()) %>%
    select(race, sex, entry, birth, exit, id, entry_age)

cohort <- init_cohort %>%
    mutate(birth = birth %>% as_date,
           entry = entry %>% as_date) %>%
    lexpand(birth, entry, exit, status = 0, breaks = list(age = c(unique(rates$start_age), Inf), per = c(unique(rates$time), Inf))) %>%
    select(age, per, id, pyrs = lex.dur) %>%
    left_join(init_cohort) %>%
    select(-exit) %>%
    as_tibble() %>%
    mutate(match_age = map_int(age, function(age) detect(rate_ages, ~ . <= age, .dir = 'backward')),
           match_per = map_int(per, function(per) detect(rate_pers, ~ . <= per, .dir = 'backward')))

cohort <- cohort %>%
    left_join(rates, by = c('match_age' = 'age', 'match_per' = 'time', 'sex', 'race')) %>%
    as_tibble()


cohort_doe <- init_cohort %>%
    mutate(birth = birth %>% as_date,
           entry = entry %>% as_date) %>%
    lexpand(birth, entry, exit, status = 0,
            breaks = list(age = c(unique(rates$start_age), Inf), per = c(unique(rates$time), Inf), fot = c(0, 10, 20, 30, 40, Inf))) %>%
    select(age, per, fot, id, pyrs = lex.dur) %>%
    left_join(init_cohort) %>%
    select(-exit) %>%
    as_tibble() %>%
    mutate(match_age = map_int(age, function(age) detect(rate_ages, ~ . <= age, .dir = 'backward')),
           match_per = map_int(per, function(per) detect(rate_pers, ~ . <= per, .dir = 'backward')))

cohort_doe <- cohort_doe %>%
    left_join(rates, by = c('match_age' = 'age', 'match_per' = 'time', 'sex', 'race')) %>%
    as_tibble()





run_sim <- function(nested_cohort, eval_times) {
    sims <- nested_cohort %>%
        mutate(sim = map_dbl(pyt, rpwe))

    o <- map_dbl(eval_times, ~observed(sims, .))
    cmr <- cmr_expected(sims, eval_times)
    smr <- smr_expected(sims, eval_times)
    ## cmr <- future_map_dbl(eval_times, ~sum(cmr_expected(sims, .)))
    ## smr <- future_map_dbl(eval_times, ~sum(smr_expected(sims, .)))

    cmr <- map2_dbl(o,  cmr, ~ .x / .y)
    smr <- map2_dbl(o, smr, ~ .x / .y)

    return(list(cmr = cmr, smr = smr, observed = o))
}

run_stratified <- function(nested_cohort, eval_times) {
    sims <- nested_cohort %>%
        mutate(sim = map_dbl(pyt, rpwe))

    outs <- stratify(sims, eval_times, mat_func = fot_matrix, lowers = c(0, 10, 20, 30, 40), uppers = c(10, 20, 30, 40, Inf))

    bind_rows(map(outs, ~mutate(as_tibble(.), eval_times = eval_times)))
}

sim_stratified <- function(cohort, eval_times, true_rate_function, nsim = 20) {
    sick_cohort <- cohort %>% true_rate_function
    nested <- sick_cohort %>% group_by(id) %>%
        nest(pyt = c(age, per, fot, pyrs, rate, start_age, end_age, match_age, match_per, true_rate)) %>%
        ungroup()
    sims <- future_map(1:nsim, ~run_stratified(nested, eval_times))
    sims
}
eval_times <- seq(from = 1951, to = 2010, by = 0.5)
eval_times_2 <- c(1990, 1995, 2000, 2005, 2010, 2015, 2020)



sim_condition <- function(cohort, eval_times, true_rate_function, nsim = 20) {
    ## change true_rate_function to take in cohort and instead
    sick_cohort <- cohort %>% true_rate_function
    nested <- sick_cohort %>% group_by(id) %>%
        nest(pyt = c(age, per, pyrs, rate, start_age, end_age, match_age, match_per, true_rate)) %>%
        ungroup()
    sims <- future_map(1:nsim, ~run_sim(nested, eval_times))
    sims
}

if (FALSE) {
baseline <- . %>% mutate(true_rate = rate)
healthy_sims <- sim_condition(cohort, eval_times, baseline, nsim = 20)

doe2 <- . %>% mutate(true_rate  = case_when(
                         fot >= 40 ~ rate * 3,
                         fot >= 30 ~ rate * 2.5,
                         fot >= 30 ~ rate * 2,
                         fot >= 40 ~ rate * 1.5,
                         TRUE ~ rate))

doe2_sims <- sim_stratified(cohort_doe, c(2010, 2020), doe2, nsim = 70)

doe2_sims %>%  bind_rows() %>% pivot_longer(c(cmr, smr)) %>% ggplot(aes(lower, value)) + geom_jitter(aes(color = name)) + facet_grid(rows = vars(eval_times))

baseline_sick <- . %>% mutate(true_rate = rate * 2)
baseline_sick <- sim_condition(cohort, eval_times, baseline_sick, nsim = 20)

sick_after_40 <- . %>% mutate(true_rate =  if_else(age > 40, rate * 2, rate))
sick_after_40 <- sim_condition(cohort, eval_times, sick_after_40, nsim = 20)

rates_double_after_1990 <- . %>% mutate(true_rate = if_else(per > 1990, rate * 2, rate))
rates_double_after_1990 <- sim_condition(cohort, eval_times, rates_double_after_1990, nsim = 20)

rates_double_after_65 <- . %>% mutate(true_rate = if_else(age > 65, rate * 2, rate))
rates_double_after_65 <- sim_condition(cohort, eval_times, rates_double_after_65, nsim = 20)


## rates increase faster with age
increase_with_age <- . %>% mutate(
                              true_rate = case_when(
                                  age >= 40 ~rate * 1.5,
                                  age >= 55 ~ rate * 2,
                                  age >= 65 ~  rate * 2.5,
                                  age >= 80 ~ rate * 3,
                                  TRUE ~ rate))
increase_with_age <- sim_condition(cohort, eval_times, increase_with_age, nsim = 20)




## rates increase faster with time
increase_with_time <- .  %>% mutate(
                              true_rate = case_when(
                                  per >= 1970~ rate * 1.5,
                                  age >= 1980~ rate * 2,
                                  age >= 1990~ rate * 2.5,
                                  age >= 2000~ rate * 3,
                                  TRUE ~ rate))
increase_with_time <- sim_condition(cohort, eval_times, increase_with_time, nsim = 20)                                     

half_triple <- . %>% mutate(true_rate = if_else(id %% 2 == 0, rate * 3, rate))
half_triple <- sim_condition(cohort, eval_times, half_triple, nsim = 20)


rates_half <- . %>% mutate(true_rate = rate / 2)
rates_half <- sim_condition(cohort, eval_times, rates_half, nsim = 20)


rates_90percent <- . %>% mutate(true_rate = rate * 0.90)
rates_90percent <- sim_condition(cohort, eval_times, rates_90percent, nsim = 20)

rates_additive <- . %>% mutate(true_rate = rate + 5/1000)
rates_additive <- sim_condition(cohort_doe, eval_times, rates_additive, nsim = 20)
}


plot_sims <- function(sims) {
    data <- map(sims, ~as_tibble(.)) %>%
        imap(~mutate(.x, id = .y, eval_times = eval_times)) %>%
        bind_rows() %>%
        select(-observed) %>%
        pivot_longer(cols = c(cmr, smr)) %>%
        mutate(unique_id = str_c(id, name))
    
    ylimmax <- data %>% filter(eval_times >= 1965) %>%
        pull(value) %>%
        max
    
    data %>% ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        theme_bw() + 
        geom_line(aes(group = unique_id)) +
        facet_zoom(xy = eval_times >= 1965, ylim = c(0, ylimmax + 0.1), horizontal = 'FALSE') 

}






## legacy stuff

   
age_stratified_comparison <- function(nested_cohort, eval_times) {
    sims <- nested_cohort %>%
        mutate(sim = map_dbl(pyt, rpwe))

    o <- observed_matrix(sims, eval_times)
    cmr <- cmr_expected_matrix(sims, eval_times)
    smr <- smr_expected_matrix(sims, eval_times)
    age <- age_matrix(sims, eval_times)

    age_20_40 <- age >= 20 & age < 40

    o_20_40 <- o
    o_20_40[] <- as.integer(if_else(age_20_40, o, FALSE))

    cmr_20_40 <- cmr
    cmr_20_40[] <- if_else(age_20_40, cmr, 0)

    smr_20_40 <- smr
    smr_20_40[] <- if_else(age_20_40, smr, 0)

    cmr_20_40 <- colSums(o_20_40) / colSums(cmr_20_40)
    smr_20_40 <- colSums(o_20_40) / colSums(smr_20_40)


        age_40_60 <- age >= 50 & age < 60

    o_40_60 <- o
    o_40_60[] <- as.integer(if_else(age_40_60, o, FALSE))

    cmr_40_60 <- cmr
    cmr_40_60[] <- if_else(age_40_60, cmr, 0)

    smr_40_60 <- smr
    smr_40_60[] <- if_else(age_40_60, smr, 0)

    cmr_40_60 <- colSums(o_40_60) / colSums(cmr_40_60)
    smr_40_60 <- colSums(o_40_60) / colSums(smr_40_60)

        age_60_80 <- age >= 60 & age < 80

    o_60_80 <- o
    o_60_80[] <- as.integer(if_else(age_60_80, o, FALSE))

    cmr_60_80 <- cmr
    cmr_60_80[] <- if_else(age_60_80, cmr, 0)

    smr_60_80 <- smr
    smr_60_80[] <- if_else(age_60_80, smr, 0)

    cmr_60_80 <- colSums(o_60_80) / colSums(cmr_60_80)
    smr_60_80 <- colSums(o_60_80) / colSums(smr_60_80)



    return(list(cmr_20_40 = cmr_20_40,
                smr_20_40 = smr_20_40,
                cmr_40_60 = cmr_40_60,
                smr_40_60 = smr_40_60,
                cmr_60_80 = cmr_60_80,
                smr_60_80 = smr_60_80))
}


age_stratified_comparison2 <- function(nested_cohort, eval_times) {
    sims <- nested_cohort %>%
        mutate(sim = map_dbl(pyt, rpwe))

    o <- observed_matrix(sims, eval_times)
    cmr <- cmr_expected_matrix(sims, eval_times)
    smr <- smr_expected_matrix(sims, eval_times)
    age <- age_matrix(sims, eval_times)

    age_20_50 <- age >= 20 & age < 50

    o_20_50 <- o
    o_20_50[] <- as.integer(if_else(age_20_50, o, FALSE))

    cmr_20_50 <- cmr
    cmr_20_50[] <- if_else(age_20_50, cmr, 0)

    smr_20_50 <- smr
    smr_20_50[] <- if_else(age_20_50, smr, 0)

    cmr_20_50 <- colSums(o_20_50) / colSums(cmr_20_50)
    smr_20_50 <- colSums(o_20_50) / colSums(smr_20_50)


    age_50p <- age >= 50 

    o_50p <- o
    o_50p[] <- as.integer(if_else(age_50p, o, FALSE))

    cmr_50p <- cmr
    cmr_50p[] <- if_else(age_50p, cmr, 0)

    smr_50p <- smr
    smr_50p[] <- if_else(age_50p, smr, 0)

    cmr_50p <- colSums(o_50p) / colSums(cmr_50p)
    smr_50p <- colSums(o_50p) / colSums(smr_50p)


    return(list(cmr_20_50 = cmr_20_50, smr_20_50 = smr_20_50, cmr_50p = cmr_50p, smr_50p = smr_50p))
}


plot_age_stratified <- function(sims) {
    data <- map(sims, ~as_tibble(.)) %>%
        imap(~mutate(.x, id = .y, eval_times = eval_times)) %>%
        bind_rows() %>%
        pivot_longer(cols = c(cmr_20_40, smr_20_40, cmr_40_60, smr_40_60, cmr_60_80, smr_60_80)) %>%
        mutate(unique_id = str_c(id, name))
    
    plot1 <- data %>% filter(name %in% c('cmr_20_40', 'smr_20_40')) %>%
        ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        theme_bw() +
        lims(y = c(0,2)) + 
        geom_line(aes(group = unique_id)) 

    plot2 <- data %>% filter(name %in% c('cmr_40_60', 'smr_40_60')) %>%
        ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        lims(y = c(0,2)) + 
        theme_bw() + 
        geom_line(aes(group = unique_id)) 
    
    plot3 <- data %>% filter(name %in% c('cmr_60_80', 'smr_60_80')) %>%
        ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        lims(y = c(0,2)) + 
        theme_bw() + 
        geom_line(aes(group = unique_id))

    list(plot1, plot2, plot3)

        ## facet_zoom(xy = eval_times >= 1965, ylim = c(0, ylimmax + 0.1), horizontal = 'FALSE') 

}


plot_age_stratified_ribbon <- function(sims) {
     map(sims, ~as_tibble(.)) %>%
        imap(~mutate(.x, id = .y, eval_times = eval_times)) %>%
        bind_rows() %>%
        pivot_longer(cols = c(cmr_20_40, smr_20_40, cmr_40_60, smr_40_60, cmr_60_80, smr_60_80)) %>%
        group_by(name, eval_times) %>%
        filter(is.finite(value), value != 0) %>%
         summarize(lower = quantile(value, 0.2), upper = quantile(value, 0.8), median = median(value)) %>%
         mutate(measure_type = str_sub(name, 1, 3)) %>%
        ggplot(aes(x = eval_times, ymin = lower, ymax = upper, y = median)) +
        geom_ribbon(aes(group = name, fill = measure_type ), alpha = 0.5) +
         geom_line(aes(group = name, color = measure_type)) +
         ylim(c(0,3)) +
         theme_bw()
}



plot_age_stratified2 <- function(sims) {
    data <- map(sims, ~as_tibble(.)) %>%
        imap(~mutate(.x, id = .y, eval_times = eval_times)) %>%
        bind_rows() %>%
        pivot_longer(cols = c(cmr_20_50, smr_20_50, cmr_50p, smr_50p)) %>%
        mutate(unique_id = str_c(id, name))
    
    plot1 <- data %>% filter(name %in% c('cmr_20_50', 'smr_20_50')) %>%
        ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        theme_bw() +
        lims(y = c(0,2)) + 
        geom_line(aes(group = unique_id)) 
    
    plot2 <- data %>% filter(name %in% c('cmr_50p', 'smr_50p')) %>%
        ggplot(aes(eval_times, value, color = name)) + geom_hline(aes(yintercept = 1), alpha = 0.3) +
        lims(y = c(0,2)) + 
        theme_bw() + 
        geom_line(aes(group = unique_id))

    list(plot1, plot2)

        ## facet_zoom(xy = eval_times >= 1965, ylim = c(0, ylimmax + 0.1), horizontal = 'FALSE') 

}


plot_age_stratified_ribbon2 <- function(sims) {
     map(sims, ~as_tibble(.)) %>%
        imap(~mutate(.x, id = .y, eval_times = eval_times)) %>%
        bind_rows() %>%
        pivot_longer(cols = c(cmr_20_50, smr_20_50, cmr_50p, smr_50p)) %>%
        group_by(name, eval_times) %>%
        filter(is.finite(value), value != 0) %>%
         summarize(lower = quantile(value, 0.2), upper = quantile(value, 0.8), median = median(value)) %>%
         mutate(measure_type = str_sub(name, 1, 3)) %>%
        ggplot(aes(x = eval_times, ymin = lower, ymax = upper, y = median)) +
        geom_ribbon(aes(group = name, fill = measure_type ), alpha = 0.5) +
         geom_line(aes(group = name, color = measure_type)) +
         ylim(c(0,3)) + theme_bw()
}


estimate.R

library(tidyverse)
library(janitor)
library(readxl)
library(mipfp)
source('table2.R')

list2env(tables, envir = environment())

start <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 110)
end   <- c(18, 28, 38, 48, 58, 68, 78, 88, 98, 108, 118)
names <- c('46-50', '51-55', '56-60', '61-65', '66-70', '71-75', '76-80', '81-85', '86-90', '91-96', '96-00')

age_groups <- c('< 5',  '5-9', '10-14', '15-19', '20-24', '25-29', '30-34', '35-39', '40-44', '45-49', '50-54', '55-59', '60-64', '65-69', '70-74', '75-79', '80-84', '85+')

col_positions <- fwf_positions(start = start, end = end, col_names = names)
## read in rates for latest time period and use only this by default
rates <- read_fwf('england_rates.txt', skip = 3, col_positions) %>%
    mutate(age_group = age_groups) %>%
    gather(time_period, rate, -age_group) %>%
    filter(time_period == '96-00') %>%
    select(-time_period)

# age group emigration numbers
age_group <- age_group %>%
    clean_names() %>%
    mutate(gender = if_else(row_number() <= 8, 'Male', 'Female')) %>%
    slice(-c(1, 9)) %>%
    gather(time_period, n, -gender_age, -gender) %>%
    rename(age = gender_age, n_age = n) %>%
    mutate(year = str_sub(time_period, 5, 8), year = parse_integer(year), time_period = NULL)

fit.extrap <- glm(n_age ~ gender + age + year, data = age_group, family = 'poisson')

# extrapolate down from 2007 down to 2000 by interpolating with poisson regression
extrap_age <- age_group %>% filter(year == 2007)

age_group_extrap <- map_dfr(c(2000:2006, 2018:2020), ~mutate(extrap_age, year = .))

pred <- broom::augment(fit.extrap, newdata = age_group_extrap, type.predict = 'response') %>%
    select(age, gender, year, n_age = .fitted) %>%
    mutate(n_age = ceiling(n_age))

emigration <- bind_rows(pred, age_group) %>% arrange(year)


## 2017 mid year estimates in northern ireland by age
pop_data <- read_excel('MYE17_SYA.xlsx', sheet = 2) %>%
    filter(
        str_detect(area,'Northern Ireland'),
        gender %in% c('Males', 'Females'),
        year %in% 2000:2017
    ) %>%
    select(age, year, MYE, gender)

## aggregate pop data into age group
pop_data <- pop_data %>%
    mutate(age_group = case_when(
               age < 18 ~ 'Less than 18 years',
               age %in% 18:24 ~ '18-24',
               age %in% 25:34 ~ '25-34',
               age %in% 35:44 ~ '35-44',
               age %in% 45:54 ~ '45-54',
               age %in% 55:64 ~ '55-64',
               age >= 65 ~ '65 years and over'
           ),
           gender = case_when(
               gender == 'Females' ~ 'Female',
               gender == 'Males' ~ 'Male'
           )) %>%
    group_by(age_group, gender, year) %>%
    summarize(n_pop = sum(MYE))

pop_data_extrap <- pop_data %>% filter(year == 2017)
pop_data_extrap <- bind_rows(
    pop_data_extrap %>% mutate(year = 2018),
    pop_data_extrap %>% mutate(year = 2019),
    pop_data_extrap %>% mutate(year = 2020))

pop_data <- pop_data %>% bind_rows(pop_data_extrap)

emigration <- emigration %>%
    rename(age_group = age) %>%
    left_join(pop_data) %>%
    mutate(rate = n_age / n_pop)


cons <- constraints
cons <- cons[ c('age_at_hire', 'time_since_first_employment', 'duration_of_employment', 'vital_status', 'sex')]

vitalc <- cons$vital_status
vitalc <- with(vitalc, list(alive = alive_confirmed + alive_assumed, dead = dead_unknown + dead_known))
cons$vital_status <- vitalc

cons <- map(cons, ~unlist(unlist(.)))

seedt <- array(1, dim = map(cons, length), dimnames = map(cons, ~names(.)))
target.list <- 1:5
weights <- Ipfp(seedt, target.list, cons)

df <- as.data.frame.table(weights[[1]]) %>% as_tibble()

## truncate replicate sample algorithm
trs <- function(table) {
    counts <- table$Freq
    rounded <- floor(counts)
    decimal <- counts - rounded
    counts <- as.integer(counts)
    topup <- sample(length(counts), round(sum(decimal)), prob = decimal)
    counts[topup] <- counts[topup] + 1
    table$integer_counts <- counts
    table
}
df <- trs(df)

cohort <- df %>% select(-Freq, -duration_of_employment) %>%
    filter(vital_status == 'alive') %>%
    select(-vital_status) %>%
    ## assume for simplicity that lowest hiring age possible is 16 and oldest is 60
    ## additionally assume that longest time since employment possible is 40 years
    mutate(
        age_at_hire = fct_recode(age_at_hire, '16-19' = '<20', '30-60' = '30+'),
        time_since_first_employment = fct_recode(time_since_first_employment, '0-20' = '<20', '30-35' = '30+')
    ) %>%
    separate(age_at_hire, into = c('aah_lb', 'aah_ub'), sep = '-') %>%
    separate(time_since_first_employment, into = c('tsfe_lb', 'tsfe_ub'), sep = '-') %>%
    mutate_at(vars(aah_lb, aah_ub, tsfe_lb, tsfe_ub), parse_integer) %>%
    mutate(
        age_lb = tsfe_lb + aah_lb,
        age_ub = tsfe_ub + aah_ub
    ) %>%
    select(age_lb, age_ub, sex, integer_counts)



## run simulations
## aggregate over sex for simplicity
emigration <- emigration %>%
    ungroup() %>%
    group_by(age_group, year) %>%
    summarize(n_age = sum(n_age),
              n_pop = sum(n_pop)) %>%
    mutate(rate = n_age / n_pop)

emigration_extreme <- filter(emigration, age_group %in% c('Less than 18 years', '65 years and over')) %>%
    ungroup() %>%
    mutate(lb = case_when(
               age_group == 'Less than 18 years' ~ 0,
               age_group == '65 years and over' ~ 65
           ),
           ub = case_when(
               age_group == 'Less than 18 years' ~ 17,
               age_group == '65 years and over' ~ Inf
           )) %>%
    select(-age_group)
emigration_normal <- filter(emigration, !(age_group %in% c('Less than 18 years', '65 years and over'))) %>%
    ungroup() %>%
    separate(age_group, into = c('lb', 'ub'), sep = '-') %>%
    mutate_at(c('lb', 'ub'), as.integer)

emigration <- bind_rows(emigration_extreme, emigration_normal)

round_to <- function(x, base) base * round(x / base)

## assume constant death rates per age group over time
## also crudely average death rates over age groups to get emigration matched groups

rates_extreme <- filter(rates, age_group %in% c('< 5', '85+')) %>%
    mutate(lb = c(0, 85), ub = c(5, Inf)) %>%
    select(-age_group)
rates_normal <- filter(rates, !(age_group %in% c('< 5', '85+'))) %>%
    separate(age_group, into = c('lb', 'ub'), sep = '-') %>%
    mutate_at(c('lb', 'ub'), as.integer)

rates <- bind_rows(rates_extreme, rates_normal)
## aggregate cohort over sex
cohort <- cohort %>%
    group_by(age_lb, age_ub) %>%
    summarize(n = sum(n)) %>%
    ungroup() %>%
    ## round to nearest 5 to simplify age group matching
    mutate(age_lb = round_to(age_lb, 5), age_ub = round_to(age_ub, 5) - 1)



age <- function(data) {
    data %>% mutate(age_lb = age_lb + 1,
                    age_ub = age_ub + 1)
}

choose_rates <- function(age_lb, age_ub) {
    rates_to_apply <- filter(rates, age_lb <= ub, lb <= age_ub)
    ## annual rate per 1k
    rate_to_apply <- mean(rates_to_apply$rate)
    rate_to_apply / 1000
}
choose_emigration <- function(age_lb, age_ub, year_cohort) {
    emigration_to_apply <- filter(emigration, year == year_cohort, age_lb <= ub, lb <= age_ub)
    emigration_rate <- sum(emigration_to_apply$n_age) / sum(emigration_to_apply$n_pop)
    emigration_rate
}
die <- function(data) {
    ## determine_rates
    local_rates <- map2_dbl(data$age_lb, data$age_ub, ~choose_rates(.x, .y))
    counts <- map_int(data$n, ~rpois(1, local_rates  * .))
    list(total_count = sum(counts), cohort = mutate(data, n = n - counts))
}

emigrate <- function(data, year) {
    local_rates <- pmap_dbl(list(data$age_lb, data$age_ub, rep(year, nrow(data))), ~choose_emigration(..1, ..2, ..3))
    counts <- map_int(data$n, ~rpois(1, local_rates * .))
    list(total_count = sum(counts), cohort = mutate(data, n = n - counts))
}

starting_cohort <- cohort

die_first <- function(starting_cohort) {
    cohort <- starting_cohort
    years <- 2000:2020
    yearly_death_counts <- length(years)
    yearly_emigration_counts <- length(years)
    ## cohort starts at 2000
    for (i in seq_along(years)) {
        ## deaths
        death_sim <- die(cohort)
        yearly_death_counts[i] <- death_sim$total_count
        cohort <- death_sim$cohort
        ## emigration
        em_sim <- emigrate(cohort, years[i])
        cohort <- em_sim$cohort
        yearly_emigration_counts[i] <- em_sim$total_count
        ## age the cohort
        cohort <- age(cohort)
    }
    list(dead = sum(yearly_death_counts), emigrated = sum(yearly_emigration_counts))
}

emigrate_first <- function(starting_cohort) {
    cohort <- starting_cohort
    years <- 2000:2020
    yearly_death_counts <- length(years)
    yearly_emigration_counts <- length(years)
    ## cohort starts at 2000
    for (i in seq_along(years)) {
        ## emigration
        em_sim <- emigrate(cohort, years[i])
        cohort <- em_sim$cohort
        yearly_emigration_counts[i] <- em_sim$total_count
        ## deaths
        death_sim <- die(cohort)
        yearly_death_counts[i] <- death_sim$total_count
        cohort <- death_sim$cohort
        ## age the cohort
        cohort <- age(cohort)
    }
    list(dead = sum(yearly_death_counts), emigrated = sum(yearly_emigration_counts))
}


